A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)

### 2018-04-13 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
updates

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flow in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domain. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all this data. This challenge gave birth to upscaling, a model order reduction method.

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition \cite{Sirovich_1987} are also MOR methods discussed in the referenced articles published in 1984 and 1987 respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed by SVD in principal component spaces.

@@

\cite{Gomes_2017} provides details on the governing equations and numerical formulation which forms the mathematical model for the multi-flow flow in this work. Albeit, it suffices to introduce only the representation of Darcy's law for immiscible multi-phase flow for the following discussion. This is written in the form:
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} =
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\alpha$ designates a phase and $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure, density and source term, which may include gravity and/or capillarity, respectively. $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$, which in turn is a function of the vector position $\mathbf{r}$, and time, $t$. Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is the absolute permeability tensor of the porous medium and it is a function of the vector position $\mathbf{r}$.

In this work, $\mathbf{K}$ is prescribed in the system of governing equations used to model multi-phase flow in porous media, this means it is an input/independent variable. Since $\mathbf{K}$ varies with position, it is impossible to determine at every point and even if it is available at every point it would require immense computer resources to utilize all the available information in a simulation. This is the reason upscaling is required for the permeability field. Upscaling can be performed from the continuum scale, across the micro scale ($mm$), local scale and meso scale ($m$) to the field scale ($Km$)\cite{ECMI_2004}. The upscaling discussed here is from the micro scale also known as the Darcy scale.

>>>>

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data. This challenge gave birth to upscaling, a model order reduction method.

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed by SVD in principal component spaces.

@@

Immiscible multi-fluid flows in porous media may be represented by the extended Darcy equation,
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} = \mathbf{u}_{\alpha}S_\alpha=
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\alpha$ designates a phase, $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity and $\mathbf{u}_{\alpha}$ is the saturation-weighted Darcy velocity. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure, density and source term, which may include gravity and/or capillarity, respectively. $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$, which in turn is a function of the vector position $\mathbf{r}$, and time, $t$.

In addition to the extended Darcy equation (Eqn.~\ref{e:darcy_eqn}), saturation conservative equations,
\begin{equation}
  \phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} + \nabla
  \cdot \left( {\mathbf u}_{\alpha} S_{\alpha}\right) =
  s_{cty,\alpha},
  \label{saturation_equation}
\end{equation}
are discretised in space with CV basis functions, and in time with the $\theta$-method \cite{pavlidis_2013b}. In Eqn.~\ref{saturation_equation}, $\phi$ is the porosity and $s_{cty}$ is a source term.

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{salinas2015,adam_2016}).

\medskip
Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is the absolute permeability tensor of the porous medium and it is a function of the vector position $\mathbf{r}$. $\mathbf{K}$ is prescribed in the system of governing conservative equations, \ie it is an indepedent input variable. Since $\mathbf{K}$ is petrophysical rock property that varies spatially, it is not possible to determine at every spatial coordinate. And, even if this information is available at every spatial coordinate, it would require immense computer resources to process during flow simulations. This is the reason upscaling is required for processing permeability variables. Upscaling can be performed from the continuum scale, across the micro scale ($mm$), local scale and meso scale ($m$) to the field scale ($km$)\cite{ECMI_2004}. The upscaling discussed here is from the micro scale also known as the Darcy scale.
